---
title: "BIOL 3991_Omics Project"
author: "Thuy Le, Sydney McLean"
date: "2026-02-25"
output: html_document
---
# Import or install the packages to use later
```{r install package, message = FALSE}
library(readxl)
library(purrr)
library(tidyverse)
library(stringr)
library(dplyr)  
library(paletteer)
```

# Sync the excel file to OneDrive
If you haven't had OneDrive on your computer yet, download it and log in by mta email.

Go to your OneDrive on your computer, there is "BIOL 3991_Omics data" folder and our excel is in it. Log in to the OneDrive on your computer and make sure you download the folder in your OneDrive by press on the clound with down arrow icon next to the folder

This is the OneDrive folder link in case you missed it:
https://mountallison-my.sharepoint.com/:f:/r/personal/ttle_mta_ca/Documents/BIOL%203991_Omics%20data?csf=1&web=1&e=DoE83U

#Import datat
```{r import data, message=FALSE}
# Find your exact file path (erase the "#" at the start of the line to use it)
## Mac OS version
hits <- list.files(path.expand("~/Library/CloudStorage"),
                   pattern = "prodatawithCELLCOUNTS\\.xlsx$",
                   recursive = TRUE,
                   full.names = TRUE)
## Window version
#hits<-list.files(path.expand("~"),
#                 pattern="prodatawithCELLCOUNTS\\.xlsx$",
#                 recursive=TRUE,
#                 full. Names=TRUE)

# You will see your file path here 
hits 

```
```{r set up data in environment}
# Paste your file path in ("")
xlsx_path <- path.expand("/Users/kyoshiinari/Library/CloudStorage/OneDrive-MountAllisonUniversity/BIOL 3991_Omics data/prodatawithCELLCOUNTS.xlsx")

sheets <- excel_sheets(xlsx_path)
raw_data <- setNames(map(sheets, ~ read_excel(xlsx_path, sheet = .x)), sheets)
print(raw_data)
```

## Figure 1: Log 10

```{r transform data and log 10 transform}
# Pick the sheet containing metabolite data
sheet_name <- names(raw_data)[purrr::map_lgl(raw_data, ~ "mtabName" %in% names(.x))][1]

df <- raw_data[[sheet_name]] %>%
  tibble::as_tibble()

# Set repl_A to NA only for these 5 metabolites 
na_replA_mets <- c(
  "glutamic acid pos", 
  "glutamine",
  "sucrose pos",
  "threonine (isom. homoserine)",
  "DMSP"
)

df <- df %>%
  mutate(
    CMP032_MIT9301_repl_A = if_else(
      mtabName %in% na_replA_mets,
      NA_real_,
      as.numeric(CMP032_MIT9301_repl_A)
    )
  )
# Identify sample columns 
sample_cols <- grep("^CMP032", names(df), value = TRUE)

# Convert to long format
long <- df %>%
  select(mtabName, all_of(sample_cols)) %>%
  pivot_longer(-mtabName, names_to = "sample", values_to = "conc") %>%
  mutate(
    conc = as.numeric(conc),
    sample_disp = sample %>%
      str_replace_all("\\.+", "_") %>%
      str_replace_all("__+", "_") %>%
      str_replace_all("_$", ""),
    Condition = str_extract(sample_disp, "repl|lown|lowp"),   # Extract the treatment group from the sample name
    Replicate = str_extract(sample_disp, "(?<=_)[A-Za-z]$"),  # Extract replicate letter (A, B, or C) from the end of the sample name
    Condition = factor(Condition, levels = c("repl", "lown", "lowp")),  # Set factor order so the samples appear in the desired order later
    Replicate = factor(Replicate, levels = c("A", "B", "C"))
  )

#  Order samples and metabolites, forces the samples to plot in the order
sample_levels <- long %>%
  distinct(sample_disp, Condition, Replicate) %>%
  arrange(Condition, Replicate) %>%
  pull(sample_disp)

met_levels <- rev(unique(df$mtabName))

# Apply the chosen factor order
long <- long %>%
  mutate(
    sample_disp = factor(sample_disp, levels = sample_levels),
    mtabName = factor(mtabName, levels = met_levels)
  )

# log10 transform 
long_log10 <- long %>%
  mutate(log10_conc = log10(conc + 1e-12))

lims_log10 <- range(long_log10$log10_conc, na.rm = TRUE)

```
 

```{r heat map log 10}
# Plot 
p_log10 <- ggplot(long_log10, aes(x = sample_disp, y = mtabName, fill = log10_conc)) +
  geom_tile() +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_viridis_c(
    option = "F",
    direction = -1,
    begin = 0.20,
    limits = lims_log10,
    breaks = c(-2.5, -5, -7.5, -10, -12),
    na.value = "grey"
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Metabolite Concentration Profiles Across Nutrient Treatments",
    x = "Treatment and Replicate",
    y = "Metabolite",
    fill = "log10(Concentration + 1e-12)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
    axis.text.y = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_log10

ggsave("metabolite_heatmap_log10.png",p_log10, width = 12, height = 8, dpi = 300)

```

# Figure 2: Log 2 Fold Change
```{r Log 2 FC data calculate}
# Set the control group and pseudocount, the pseudocount prevents division by zero when concentrations are 0
control_group <- "repl"
pseudo <- 1e-12

# Calculated log2 fc
fc_long <- long %>%
  group_by(mtabName) %>%
  mutate(
    control_mean = mean(conc[Condition == control_group], na.rm = TRUE),
    log2FC = log2((conc + pseudo) / (control_mean + pseudo))
  ) %>%
  ungroup()

```


Sydney ver 
df_log2fc <- df %>%

  mutate(control_mean = rowMeans(select(., contains("repl")), na.rm = TRUE)) %>%

  mutate(across(starts_with("CMP"), ~ log2((. + 1e-12) / (control_mean + 1e-12)))) %>%

  select(mtabName, starts_with("CMP")) %>%

  pivot_longer(-mtabName, names_to = "Sample", values_to = "Log2FC")
  
ggplot(df_log2fc, aes(x = Sample, y = mtabName, fill = Log2FC)) +
  geom_tile(color = "white", linewidth = 0.1) +
  geom_vline(xintercept = c(3.5, 6.5), color = "black", linewidth = 0.5) +
  scale_fill_gradient2(low = "pink", mid = "white", high = "red", 
                       midpoint = 0, limit = c(-8, 8), oob = scales::squish) +
  theme_minimal() +
  labs(title = "Differential Metabolic Profiles",
       subtitle = "Log2 Fold Change relative to Control (repl) mean",
       x = "Samples", y = "Metabolites") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 7))

```{r Log 2 FC normal map}
# Plot (set limit from -5.0 to 5.0)
max_abs <- max(abs(fc_long$log2FC), na.rm = TRUE)

p_fc <- ggplot(fc_long, aes(x = sample_disp, y = mtabName, fill = log2FC)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_distiller(
  palette = "RdBu",
    direction = -1,
    limit = c(-5.0, 5.0), oob = scales::squish,
    na.value = "grey"
) +
  labs(fill = "log2FC") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Different Metabolic Profiles",
    subtitle = "Log2 Fold Change relative to control group mean",
    x = "Treatment and Replicate",
    y = "Metabolites",
    fill = "log2FC"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12, , color = "black"),
    axis.text.y = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_fc

ggsave("metabolite_heatmap_log2fc.png", p_fc, width = 10, height = 8, dpi = 300)

# Plot (same as Sydney setting)
p_fc2 <- ggplot(fc_long, aes(x = sample_disp, y = mtabName, fill = log2FC)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
 geom_vline(xintercept = c(3.5, 6.5), color = "black", linewidth = 0.5) +
  scale_fill_gradient2(low = "pink", mid = "white", high = "red", 
                       midpoint = 0, limit = c(-5.0, 5.0), oob = scales::squish, na.value = "grey"
) +
  labs(fill = "log2FC") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Different Metabolic Profiles",
    subtitle = "Log2 Fold Change relative to control group mean",
    x = "Treatment and Replicate",
    y = "Metabolites",
    fill = "log2FC"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12, , color = "black"),
    axis.text.y = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_fc2

# Plot (set limit as normal)
p_fc3 <- ggplot(fc_long, aes(x = sample_disp, y = mtabName, fill = log2FC)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_distiller(
  palette = "RdBu",
    direction = -1,
    limit = c(-max_abs,max_abs), oob = scales::squish,
    na.value = "grey"
) +
  labs(fill = "log2FC") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Different Metabolic Profiles",
    subtitle = "Log2 Fold Change relative to control group mean",
    x = "Treatment and Replicate",
    y = "Metabolites",
    fill = "log2FC"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12, , color = "black"),
    axis.text.y = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_fc3


```

```{r Column-normalized data}
# Build a numeric matrix from the sample columns
mat <- df %>%
  select(all_of(sample_cols)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

# Normalize across columns. This standardizes each sample column separately
col_z <- scale(mat, center = TRUE, scale = TRUE)
# If a column has no variation, scale() can produce Na. Convert those to NA so they plot as blank
col_z[is.nan(col_z)] <- NA  

# Convert back to long format
col_z_long <- as_tibble(as.data.frame(col_z, check.names = FALSE)) %>%
  mutate(mtabName = df$mtabName, .before = 1) %>%
  pivot_longer(-mtabName, names_to = "sample", values_to = "z_score") %>%
  mutate(
    sample_disp = sample %>%
      str_replace_all("\\.+", "_") %>%
      str_replace_all("__+", "_") %>%
      str_replace_all("_$", ""),
    Condition = str_extract(sample_disp, "repl|lown|lowp"),
    Replicate = str_extract(sample_disp, "(?<=_)[A-Za-z]$"),
    Condition = factor(Condition, levels = c("repl", "lown", "lowp")),
    Replicate = factor(Replicate, levels = c("A", "B", "C")),
    sample_disp = factor(sample_disp, levels = sample_levels),
    mtabName = factor(mtabName, levels = met_levels)
  )
```

```{r Log 2 FC column-normalized heatmap}
#
max_abs_col <- max(abs(col_z_long$z_score), na.rm = TRUE)

# Plot heatmap
p_col <- ggplot(col_z_long, aes(x = sample_disp, y = mtabName, fill = z_score)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_distiller(
    palette = "RdBu",
    direction = -1,
    limits = c(-max_abs_col, max_abs_col),
    na.value = "white"
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Column-normalized Metabolite Heatmap",
    x = "Treatment and Replicate",
    y = "Metabolite",
    fill = "Column z-score"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
    axis.text.y = element_text(size = 11),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_col

ggsave("metabolite_heatmap_column_normalized.png", p_col, width = 10, height = 8, dpi = 300)
```

```{r Row-normalize data}
#
row_z <- t(scale(t(mat), center = TRUE, scale = TRUE))
row_z[is.nan(row_z)] <- NA

#
row_z_long <- as_tibble(as.data.frame(row_z, check.names = FALSE)) %>%
  mutate(mtabName = df$mtabName, .before = 1) %>%
  pivot_longer(-mtabName, names_to = "sample", values_to = "z_score") %>%
  mutate(
    sample_disp = sample %>%
      str_replace_all("\\.+", "_") %>%
      str_replace_all("__+", "_") %>%
      str_replace_all("_$", ""),
    Condition = str_extract(sample_disp, "repl|lown|lowp"),
    Replicate = str_extract(sample_disp, "(?<=_)[A-Za-z]$"),
    Condition = factor(Condition, levels = c("repl", "lown", "lowp")),
    Replicate = factor(Replicate, levels = c("A", "B", "C")),
    sample_disp = factor(sample_disp, levels = sample_levels),
    mtabName = factor(mtabName, levels = met_levels)
  )
```

```{r Log 2 FC row-normalize heatmap}
#
max_abs_row <- max(abs(row_z_long$z_score), na.rm = TRUE)

# Plot heatmap
p_row <- ggplot(row_z_long, aes(x = sample_disp, y = mtabName, fill = z_score)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_distiller(
    palette = "RdBu",
    direction = -1,
    limits = c(-max_abs_row, max_abs_row),
    na.value = "white"
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Row-normalized Metabolite Heatmap",
    x = "Treatment and Replicate",
    y = "Metabolite",
    fill = "Row z-score"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
    axis.text.y = element_text(size = 11),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_row

ggsave("metabolite_heatmap_row_normalized.png", p_row, width = 10, height = 8, dpi = 300)
```

