---
title: "BIOL 3991_Omics Project"
author: "Thuy Le"
date: "2026-02-25"
output: html_document
---
# Import or install the packages to use later
```{r install package}
library(readxl)
library(purrr)
library(tidyverse)
```

# Sync the excel file to OneDrive
If you haven't had OneDrive on your computer yet, download it and log in by mta email.

Go to your OneDrive on your computer, there is "BIOL 3991_Omics data" folder and our excel is in it. Log in to the OneDrive on your computer and make sure you download the folder in your OneDrive by press on the clound with down arrow icon next to the folder

#Import datat
```{r import data, message=FALSE}
# Find your exact file path
hits <- list.files(path.expand("~/Library/CloudStorage"),
                   pattern = "prodataMULTIPLIED\\.xlsx$",
                   recursive = TRUE,
                   full.names = TRUE)
hits

# Paste your file path in ""
xlsx_path <- path.expand("/Users/kyoshiinari/Library/CloudStorage/OneDrive-MountAllisonUniversity/BIOL 3991_Omics data/prodataMULTIPLIED.xlsx")

sheets <- excel_sheets(xlsx_path)
raw_data <- setNames(map(sheets, ~ read_excel(xlsx_path, sheet = .x)), sheets)
print(raw_data)
```

## Including Plots

You can also embed plots, for example:

```{r }
plot(pressure)
```








