---
title: "BIOL 3991_Omics Project"
author: "Thuy Le, Sydney McLean"
date: "2026-02-25"
output: html_document
---
# Import or install the packages to use later
```{r install package, message = FALSE}
library(readxl)
library(purrr)
library(tidyverse)
library(stringr)
library(grid)
library(dplyr)  
library(paletteer)
library(pheatmap)
```

# Sync the excel file to OneDrive
If you haven't had OneDrive on your computer yet, download it and log in by mta email.

Go to your OneDrive on your computer, there is "BIOL 3991_Omics data" folder and our excel is in it. Log in to the OneDrive on your computer and make sure you download the folder in your OneDrive by press on the clound with down arrow icon next to the folder

This is the OneDrive folder link in case you missed it:
https://mountallison-my.sharepoint.com/:f:/r/personal/ttle_mta_ca/Documents/BIOL%203991_Omics%20data?csf=1&web=1&e=DoE83U

#Import datat
```{r import data, message=FALSE}
# Find your exact file path (erase the "#" at the start of the line to use it)
## Mac OS version
hits <- list.files(path.expand("~/Library/CloudStorage"),
                   pattern = "prodatawithCELLCOUNTS\\.xlsx$",
                   recursive = TRUE,
                   full.names = TRUE)
## Window version
#hits<-list.files(path.expand("~"),
#                 pattern="prodatawithCELLCOUNTS\\.xlsx$",
#                 recursive=TRUE,
#                 full. Names=TRUE)

# You will see your file path here 
hits 

```
```{r set up data in environment}
# Paste your file path in ("")
xlsx_path <- path.expand("/Users/kyoshiinari/Library/CloudStorage/OneDrive-MountAllisonUniversity/BIOL 3991_Omics data/prodatawithCELLCOUNTS.xlsx")

sheets <- excel_sheets(xlsx_path)
raw_data <- setNames(map(sheets, ~ read_excel(xlsx_path, sheet = .x)), sheets)
print(raw_data)
```

## Figure 1: Log 10

```{r transform data and log 10 transform}
# Pick the sheet containing metabolite data
sheet_name <- names(raw_data)[purrr::map_lgl(raw_data, ~ "mtabName" %in% names(.x))][1]

df <- raw_data[[sheet_name]] %>%
  tibble::as_tibble()

# Set repl_A to NA only for these 5 metabolites 
na_replA_mets <- c(
  "glutamic acid pos", 
  "glutamine",
  "sucrose pos",
  "threonine (isom. homoserine)",
  "DMSP"
)

df <- df %>% 
  mutate( CMP032_MIT9301_repl_A = if_else ( 
          mtabName %in% na_replA_mets, 
          NA_real_, 
          as.numeric(CMP032_MIT9301_repl_A) 
          ) 
        )
# Identify sample columns 
sample_cols <- grep("^CMP032", names(df), value = TRUE)

# Convert to long format
long <- df %>%
  select(mtabName, all_of(sample_cols)) %>%
  pivot_longer(-mtabName, names_to = "sample", values_to = "conc") %>%
  mutate(
    conc = as.numeric(conc),
    sample_disp = sample %>%
      str_replace_all("\\.+", "_") %>%
      str_replace_all("__+", "_") %>%
      str_replace_all("_$", ""),
    Condition = str_extract(sample_disp, "repl|lown|lowp"),   # Extract the treatment group from the sample name
    Replicate = str_extract(sample_disp, "(?<=_)[A-Za-z]$"),  # Extract replicate letter (A, B, or C) from the end of the sample name
    Condition = factor(Condition, levels = c("repl", "lown", "lowp")),  # Set factor order so the samples appear in the desired order later
    Replicate = factor(Replicate, levels = c("A", "B", "C"))
  )

#  Order samples and metabolites, forces the samples to plot in the order
sample_levels <- long %>%
  distinct(sample_disp, Condition, Replicate) %>%
  arrange(Condition, Replicate) %>%
  pull(sample_disp)

met_levels <- rev(unique(df$mtabName))

# Apply the chosen factor order
long <- long %>%
  mutate(
    sample_disp = factor(sample_disp, levels = sample_levels),
    mtabName = factor(mtabName, levels = met_levels)
  )

# log10 transform 
long_log10 <- long %>%
  mutate(log10_conc = log10(conc + 1e-12))

lims_log10 <- range(long_log10$log10_conc, na.rm = TRUE)

```
 

```{r heat map log 10}
# Plot 
p_log10 <- ggplot(long_log10, aes(x = sample_disp, y = mtabName, fill = log10_conc)) +
  geom_tile() +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_viridis_c(
    option = "F",
    direction = -1,
    begin = 0.20,
    limits = lims_log10,
    breaks = c(-2.5, -5, -7.5, -10, -12),
    na.value = "grey"
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Metabolite Concentration Profiles Across Nutrient Treatments",
    x = "Treatment and Replicate",
    y = "Metabolite",
    fill = "log10(Concentration + 1e-12)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
    axis.text.y = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_log10

ggsave("metabolite_heatmap_log10.png",p_log10, width = 12, height = 8, dpi = 300)

```

# Figure 2: Log 2 Fold Change
```{r Log 2 FC data calculate}
# Set the control group and pseudocount, the pseudocount prevents division by zero when concentrations are 0
control_group <- "repl"
pseudo <- 1e-12

# Calculated log2 fc
fc_long <- long %>%
  group_by(mtabName) %>%
  mutate(
    control_mean = mean(conc[Condition == control_group], na.rm = TRUE),
    log2FC = log2((conc + pseudo) / (control_mean + pseudo))
  ) %>%
  ungroup()

```


Sydney ver 
df_log2fc <- df %>%

  mutate(control_mean = rowMeans(select(., contains("repl")), na.rm = TRUE)) %>%

  mutate(across(starts_with("CMP"), ~ log2((. + 1e-12) / (control_mean + 1e-12)))) %>%

  select(mtabName, starts_with("CMP")) %>%

  pivot_longer(-mtabName, names_to = "Sample", values_to = "Log2FC")
  
ggplot(df_log2fc, aes(x = Sample, y = mtabName, fill = Log2FC)) +
  geom_tile(color = "white", linewidth = 0.1) +
  geom_vline(xintercept = c(3.5, 6.5), color = "black", linewidth = 0.5) +
  scale_fill_gradient2(low = "pink", mid = "white", high = "red", 
                       midpoint = 0, limit = c(-8, 8), oob = scales::squish) +
  theme_minimal() +
  labs(title = "Differential Metabolic Profiles",
       subtitle = "Log2 Fold Change relative to Control (repl) mean",
       x = "Samples", y = "Metabolites") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 7))

```{r Log 2 FC normal map}
# Plot (set limit from -5.0 to 5.0)
max_abs <- max(abs(fc_long$log2FC), na.rm = TRUE)

p_fc <- ggplot(fc_long, aes(x = sample_disp, y = mtabName, fill = log2FC)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_distiller(
  palette = "RdBu",
    direction = -1,
    limit = c(-5.0, 5.0), oob = scales::squish,
    na.value = "grey"
) +
  labs(fill = "log2FC") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Different Metabolic Profiles",
    subtitle = "Log2 Fold Change relative to control group mean",
    x = "Treatment and Replicate",
    y = "Metabolites",
    fill = "log2FC"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12, , color = "black"),
    axis.text.y = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_fc

ggsave("metabolite_heatmap_log2fc.png", p_fc, width = 10, height = 8, dpi = 300)

# Plot (same as Sydney setting)
p_fc2 <- ggplot(fc_long, aes(x = sample_disp, y = mtabName, fill = log2FC)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
 geom_vline(xintercept = c(3.5, 6.5), color = "black", linewidth = 0.5) +
  scale_fill_gradient2(low = "pink", mid = "white", high = "red", 
                       midpoint = 0, limit = c(-5.0, 5.0), oob = scales::squish, na.value = "grey"
) +
  labs(fill = "log2FC") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Different Metabolic Profiles",
    subtitle = "Log2 Fold Change relative to control group mean",
    x = "Treatment and Replicate",
    y = "Metabolites",
    fill = "log2FC"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12, , color = "black"),
    axis.text.y = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_fc2

# Plot (set limit as normal)
p_fc3 <- ggplot(fc_long, aes(x = sample_disp, y = mtabName, fill = log2FC)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_distiller(
  palette = "RdBu",
    direction = -1,
    limit = c(-max_abs,max_abs), oob = scales::squish,
    na.value = "grey"
) +
  labs(fill = "log2FC") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Different Metabolic Profiles",
    subtitle = "Log2 Fold Change relative to control group mean",
    x = "Treatment and Replicate",
    y = "Metabolites",
    fill = "log2FC"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12, , color = "black"),
    axis.text.y = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_fc3


```

```{r Column-normalized data}
# Build a numeric matrix from the sample columns
mat <- df %>%
  select(all_of(sample_cols)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

# Normalize across columns. This standardizes each sample column separately
col_z <- scale(mat, center = TRUE, scale = TRUE)
# If a column has no variation, scale() can produce Na. Convert those to NA so they plot as blank
col_z[is.nan(col_z)] <- NA  

# Convert back to long format
col_z_long <- as_tibble(as.data.frame(col_z, check.names = FALSE)) %>%
  mutate(mtabName = df$mtabName, .before = 1) %>%
  pivot_longer(-mtabName, names_to = "sample", values_to = "z_score") %>%
  mutate(
    sample_disp = sample %>%
      str_replace_all("\\.+", "_") %>%
      str_replace_all("__+", "_") %>%
      str_replace_all("_$", ""),
    Condition = str_extract(sample_disp, "repl|lown|lowp"),
    Replicate = str_extract(sample_disp, "(?<=_)[A-Za-z]$"),
    Condition = factor(Condition, levels = c("repl", "lown", "lowp")),
    Replicate = factor(Replicate, levels = c("A", "B", "C")),
    sample_disp = factor(sample_disp, levels = sample_levels),
    mtabName = factor(mtabName, levels = met_levels)
  )
```

```{r Log 2 FC column-normalized heatmap}
#
max_abs_col <- max(abs(col_z_long$z_score), na.rm = TRUE)

# Plot heatmap
p_col <- ggplot(col_z_long, aes(x = sample_disp, y = mtabName, fill = z_score)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_distiller(
    palette = "RdBu",
    direction = -1,
#    limits = c(-max_abs_col, max_abs_col),
    limit = c(-1.7, 1.7), oob = scales::squish,
    na.value = "grey"
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Column-normalized Metabolite Heatmap",
    x = "Treatment and Replicate",
    y = "Metabolite",
    fill = "z-score"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
    axis.text.y = element_text(size = 11),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_col

ggsave("metabolite_heatmap_column_normalized.png", p_col, width = 10, height = 8, dpi = 300)
```

```{r Row-normalize data}
#
row_z <- t(scale(t(mat), center = TRUE, scale = TRUE))
row_z[is.nan(row_z)] <- NA

#
row_z_long <- as_tibble(as.data.frame(row_z, check.names = FALSE)) %>%
  mutate(mtabName = df$mtabName, .before = 1) %>%
  pivot_longer(-mtabName, names_to = "sample", values_to = "z_score") %>%
  mutate(
    sample_disp = sample %>%
      str_replace_all("\\.+", "_") %>%
      str_replace_all("__+", "_") %>%
      str_replace_all("_$", ""),
    Condition = str_extract(sample_disp, "repl|lown|lowp"),
    Replicate = str_extract(sample_disp, "(?<=_)[A-Za-z]$"),
    Condition = factor(Condition, levels = c("repl", "lown", "lowp")),
    Replicate = factor(Replicate, levels = c("A", "B", "C")),
    sample_disp = factor(sample_disp, levels = sample_levels),
    mtabName = factor(mtabName, levels = met_levels)
  )
```

```{r Log 2 FC row-normalize heatmap}
#
max_abs_row <- max(abs(row_z_long$z_score), na.rm = TRUE)

# Plot heatmap
p_row <- ggplot(row_z_long, aes(x = sample_disp, y = mtabName, fill = z_score)) +
  geom_tile(color = "white", linewidth = 0.25) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_distiller(
    palette = "RdBu",
    direction = -1,
    limits = c(-max_abs_row, max_abs_row),
    na.value = "grey"
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Row-normalized Metabolite Heatmap",
    x = "Treatment and Replicate",
    y = "Metabolite",
    fill = "z-score"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
    axis.text.y = element_text(size = 11),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_row

ggsave("metabolite_heatmap_row_normalized.png", p_row, width = 10, height = 8, dpi = 300)
```

# Heat map + dendrogram:

```{r Row normalize Heatmap with dendrogram add data making}
# Prepare row-normalized matrix for dendrogram heatmap

# Build numeric matrix from the sample columns
mat_dend <- df %>%
  select(all_of(sample_cols)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

# Use metabolite names as row names
rownames(mat_dend) <- df$mtabName

# Clean sample names so they match the sample order used earlier
colnames(mat_dend) <- colnames(mat_dend) %>%
  str_replace_all("\\.+", "_") %>%
  str_replace_all("__+", "_") %>%
  str_replace_all("_$", "")

# Keep columns in biological order: repl, lown, lowp
mat_dend <- mat_dend[, sample_levels, drop = FALSE]


# Row-normalize: z-score each metabolite across samples
row_z_display <- t(scale(t(mat_dend), center = TRUE, scale = TRUE))

# Keep NA values in the display matrix so they plot as gray
row_z_display[is.nan(row_z_display)] <- NA

# Make a second copy only for clustering
# Replace NA with 0 here so hierarchical clustering can run
row_z_for_cluster <- row_z_display
row_z_for_cluster[is.na(row_z_for_cluster)] <- 0

# Build the row dendrogram
row_hc <- hclust(dist(row_z_for_cluster), method = "complete")

# Clip z-scores for plotting so the colors are easier to read
row_z_plot <- row_z_display
row_z_plot[row_z_plot > 2] <- 2
row_z_plot[row_z_plot < -2] <- -2
```

```{r Row normalize heat map + dendrogram, fig.width = 12, fig.height = 8}
# Column annotations
annotation_col <- data.frame(
  Treatment = c("control", "control", "control", "low nitrogen", "low nitrogen", "low nitrogen", "low phosphorus", "low phosphorus", "low phosphorus"),
  Replicate = c("A", "B", "C", "A", "B", "C", "A", "B", "C")
)

rownames(annotation_col) <- sample_levels

# Choose cleaner annotation colors
annotation_colors <- list(
  Treatment = c(
    "control" = "#7AA6DC",
    "low nitrogen" = "#C77CFF",
    "low phosphorus" = "#F4A582"
  ),
  Replicate = c(
    A = "#81A665",
    B = "#FFD92F",
    C = "#E87687"
  )
)

# Use the same RdBu-style palette as your ggplot heatmaps
hm_cols <- colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100)
hm_breaks <- seq(-2, 2, length.out = 101)

png("row-normalized_heatmap_dendrogram.png" , width = 2400, height = 1600, res = 170)

pheatmap::pheatmap (
  row_z_plot,
  color = hm_cols,
  breaks = hm_breaks,
  cluster_rows = row_hc,
  cluster_cols = FALSE,
  gaps_col = c(3, 6),
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  na_col = "grey",
  border_color = "white",
  fontsize_row = 10,
  fontsize_col = 10,
  angle_col = "45",
  cellheight = 15,
  cellwidth = 30,
  treeheight_row = 110,
  treeheight_col = 0,
  legend_breaks = c(-2, -1, 0, 1, 2),
  legend_labels = c("-2", "-1", "0", "1", "2"),
  main = "Row-normalized Metabolite Heatmap with Dendrogram"
)
dev.off()
```

```{r Log 10 heatmap + dendrogram data }
# ---- Build wide matrix for log10 heatmap with its own dendrogram ----

# Start from the sample columns only
mat_log10 <- df %>%
  select(all_of(sample_cols)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

# Add metabolite names as row names
rownames(mat_log10) <- df$mtabName

# Clean sample names so they match your plotting order
colnames(mat_log10) <- colnames(mat_log10) %>%
  str_replace_all("\\.+", "_") %>%
  str_replace_all("__+", "_") %>%
  str_replace_all("_$", "")

# Keep columns in treatment order
mat_log10 <- mat_log10[, sample_levels, drop = FALSE]

# Log10 transform
# 0 stays in the data and becomes log10(0 + 1e-12) = -12
# NA stays NA
log10_display <- log10(mat_log10 + 1e-12)

# Make a second version only for clustering
# Replace NA with the row mean so the dendrogram can still be built
log10_for_cluster <- log10_display

for (i in seq_len(nrow(log10_for_cluster))) {
  row_vals <- log10_for_cluster[i, ]
  row_mean <- mean(row_vals, na.rm = TRUE)
  if (is.nan(row_mean)) row_mean <- -12
  row_vals[is.na(row_vals)] <- row_mean
  log10_for_cluster[i, ] <- row_vals
}

# Build a log10-specific row dendrogram
row_hc_log10 <- hclust(dist(log10_for_cluster), method = "complete")

```

```{r Log 10 heatmap +dendrogram plot, fig.width = 12, fig.height = 8}
# Viridis-like palette for log10
log10_cols <- viridisLite::viridis(100, option = "F", direction = -1, begin = 0.20)
log10_breaks <- seq(min(log10_display, na.rm = TRUE), -2, length.out = 101)

png("metabolite_heatmap_log10_dendrogram.png", width = 2600, height = 1800, res = 180)

pheatmap::pheatmap(
  log10_display,
  color = log10_cols,
  breaks = log10_breaks,
  cluster_rows = row_hc_log10,
  cluster_cols = FALSE,
  gaps_col = c(3, 6),
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  na_col = "grey",
  border_color = "white",
  fontsize_row = 10,
  fontsize_col = 10,
  angle_col = "45",
  cellheight = 15,
  cellwidth = 30,
  treeheight_row = 150,
  treeheight_col = 0,
  main = "Log10 Metabolite Heatmap"
)

dev.off()
```

```{r log 2 fc heatmap data + dendrogram data}
# Build wide matrix for log2FC heatmap with its own dendrogram

fc_mat <- fc_long %>%
  select(mtabName, sample_disp, log2FC) %>%
  mutate(sample_disp = as.character(sample_disp)) %>%
  pivot_wider(names_from = sample_disp, values_from = log2FC) %>%
  tibble::column_to_rownames("mtabName") %>%
  as.matrix()

# Keep columns in biological order
fc_mat <- fc_mat[, sample_levels, drop = FALSE]

# Matrix for plotting
fc_plot <- fc_mat

# Make a second version only for clustering
# Replace NA with the row mean; if the whole row is NA, use 0
fc_for_cluster <- fc_mat

for (i in seq_len(nrow(fc_for_cluster))) {
  row_vals <- fc_for_cluster[i, ]
  row_mean <- mean(row_vals, na.rm = TRUE)
  if (is.nan(row_mean)) row_mean <- 0
  row_vals[is.na(row_vals)] <- row_mean
  fc_for_cluster[i, ] <- row_vals
}

# Build a log2FC-specific row dendrogram
row_hc_fc <- hclust(dist(fc_for_cluster), method = "complete")

# Use full data range
max_abs_fc <- max(abs(fc_plot), na.rm = TRUE)
fc_cols <- colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100)
fc_breaks <- seq(-max_abs_fc, max_abs_fc, length.out = 101)
```

```{r log 2 fc heatmap data + dendrogram plotting}
png("log2fc_heatmap_dendrogram.png", width = 2400, height = 1600, res = 170)

pheatmap::pheatmap(
  fc_plot,
  color = fc_cols,
  breaks = fc_breaks,
  cluster_rows = row_hc_fc,
  cluster_cols = FALSE,
  gaps_col = c(3, 6),
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  na_col = "grey",
  border_color = "white",
  fontsize_row = 10,
  fontsize_col = 10,
  angle_col = "45",
  cellheight = 15,
  cellwidth = 30,
  treeheight_row = 110,
  treeheight_col = 0,
  legend_breaks = c(-max_abs_fc, 0, max_abs_fc),
  legend_labels = round(c(-max_abs_fc, 0, max_abs_fc), 2),
  main = "Log2 Fold Change Heatmap with Dendrogram"
)

dev.off()
```

```{r column normalize heatmap + dendrogram data & plotting}
# ---- Build wide matrix for column-normalized heatmap with its own dendrogram ----

mat_col <- df %>%
  select(all_of(sample_cols)) %>%
  mutate(across(everything(), as.numeric)) %>%
  as.matrix()

rownames(mat_col) <- df$mtabName

colnames(mat_col) <- colnames(mat_col) %>%
  str_replace_all("\\.+", "_") %>%
  str_replace_all("__+", "_") %>%
  str_replace_all("_$", "")

# Keep columns in biological order
mat_col <- mat_col[, sample_levels, drop = FALSE]

# Column-normalize
col_z_display <- scale(mat_col, center = TRUE, scale = TRUE)
col_z_display[is.nan(col_z_display)] <- NA

# Make a second version only for clustering
# Replace NA with the row mean; if the whole row is NA, use 0
col_z_for_cluster <- col_z_display

for (i in seq_len(nrow(col_z_for_cluster))) {
  row_vals <- col_z_for_cluster[i, ]
  row_mean <- mean(row_vals, na.rm = TRUE)
  if (is.nan(row_mean)) row_mean <- 0
  row_vals[is.na(row_vals)] <- row_mean
  col_z_for_cluster[i, ] <- row_vals
}

# Build a column-normalized-specific row dendrogram
row_hc_col <- hclust(dist(col_z_for_cluster), method = "complete")

# Keep original values for plotting
col_z_plot <- col_z_display

# Use the full display range
max_abs_col <- max(abs(col_z_plot), na.rm = TRUE)
col_cols <- colorRampPalette(rev(RColorBrewer::brewer.pal(11, "RdBu")))(100)
col_breaks <- seq(-max_abs_col, max_abs_col, length.out = 101)

png("column_normalized_heatmap_dendrogram.png", width = 2400, height = 1600, res = 170)

pheatmap::pheatmap(
  col_z_plot,
  color = col_cols,
  breaks = col_breaks,
  cluster_rows = row_hc_col,
  cluster_cols = FALSE,
  gaps_col = c(3, 6),
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  na_col = "grey",
  border_color = "white",
  fontsize_row = 10,
  fontsize_col = 10,
  angle_col = "45",
  cellheight = 15,
  cellwidth = 30,
  treeheight_row = 150,
  treeheight_col = 0,
  legend_breaks = c(-max_abs_col, 0, max_abs_col),
  legend_labels = round(c(-max_abs_col, 0, max_abs_col), 2),
  main = "Column-normalized Metabolite Heatmap with Dendrogram"
)

dev.off()
```

# Top 10 metabolite

```{r Top 10 metabolite data & plotting, message = FALSE}
# ---- Figure 5: Top 10 rank-abundance plots, stacked vertically ----

# Summarize mean concentration for each metabolite within each treatment
rank_df <- long %>%
  group_by(Condition, mtabName) %>%
  summarise(
    mean_conc = mean(conc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Treatment = case_when(
      Condition == "repl" ~ "Control",
      Condition == "lown" ~ "Low nitrogen",
      Condition == "lowp" ~ "Low phosphorus"
    )
  )

# Find the top 10 metabolites within each treatment
top10_mets <- rank_df %>%
  group_by(Treatment) %>%
  slice_max(order_by = mean_conc, n = 10, with_ties = FALSE) %>%
  arrange(desc(mean_conc), .by_group = TRUE) %>%
  select(Treatment, mtabName)

# Go back to the original long data and keep all 3 replicates
plot_df <- long %>%
  mutate(
    Treatment = case_when(
      Condition == "repl" ~ "Control",
      Condition == "lown" ~ "Low nitrogen",
      Condition == "lowp" ~ "Low phosphorus"
    )
  ) %>%
  inner_join(top10_mets, by = c("Treatment", "mtabName"))

# Create the metabolite order within each treatment
met_order <- rank_df %>%
  inner_join(top10_mets, by = c("Treatment", "mtabName")) %>%
  group_by(Treatment) %>%
  arrange(desc(mean_conc), .by_group = TRUE) %>%
  mutate(
    metab_facet = paste(Treatment, mtabName, sep = "__")
  ) %>%
  ungroup() %>%
  distinct(Treatment, mtabName, metab_facet)

plot_df <- plot_df %>%
  left_join(met_order, by = c("Treatment", "mtabName")) %>%
  mutate(
    metab_facet = factor(metab_facet, levels = unique(met_order$metab_facet)),
    Replicate = factor(Replicate, levels = c("A", "B", "C")),
    conc_plot = conc + 1e-12   # needed so 0 values can appear on a log scale
  )

plot_df_no_outlier <- plot_df %>%
  mutate(log10_conc = log10(conc_plot)) %>%
  group_by(Treatment) %>%
  mutate(
    Q1 = quantile(log10_conc, 0.25, na.rm = TRUE),
    Q3 = quantile(log10_conc, 0.75, na.rm = TRUE),
    IQR_val = Q3 - Q1,
    lower_bound = Q1 - 1.5 * IQR_val,
    upper_bound = Q3 + 1.5 * IQR_val,
    keep_point = log10_conc >= lower_bound & log10_conc <= upper_bound
  ) %>%
  ungroup() %>%
  filter(keep_point)

# Plot as 3 stacked panels
p_rank_top10_log <- ggplot(plot_df_no_outlier, aes(x = metab_facet, y = conc_plot, color = Replicate)) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.8, na.rm = TRUE)  +
  facet_wrap(~ Treatment, ncol = 1, scales = "free_x") +
  scale_x_discrete(labels = function(x) sub("^.*__", "", x)) +
  scale_y_log10() +
  scale_color_manual(values = c("A" = "#0072B2", "B" = "#D55E00", "C" = "#009E73")) +
  labs(
    title = "Top 10 Most Abundant Metabolites Concentrations",
    x = "Metabolite",
    y = "Concentration (log)",
    color = "Replicate"
  ) +
  theme_bw(12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    strip.text = element_text(face = "bold", size = 13),
    panel.spacing = unit(1.2, "lines")
  )


p_rank_top10_log

ggsave("log_top10_rank_abundance_by_treatment.png", p_rank_top10_log,
       width = 11, height = 10, dpi = 300)
```

```{r Top 10 metabolite tranform log data & plotting, message = FALSE}
top10_rank_df_log <- rank_df %>%
  mutate(mean_conc_plot = mean_conc + 1e-12) %>%
  group_by(Treatment) %>%
  slice_max(order_by = mean_conc, n = 10, with_ties = FALSE) %>%
  arrange(desc(mean_conc), .by_group = TRUE) %>%
  mutate(
    metab_facet = paste(Treatment, mtabName, sep = "__"),
    metab_facet = factor(metab_facet, levels = unique(metab_facet))
  ) %>%
  ungroup()

p_rank_top10_log <- ggplot(top10_rank_df_log, aes(x = metab_facet, y = mean_conc_plot, group = 1)) +
#  geom_line(color = "steelblue") +
  geom_point(color = "steelblue", size = 2.5) +
  facet_wrap(~ Treatment, ncol = 1, scales = "free_x") +
  scale_x_discrete(labels = function(x) sub("^.*__", "", x)) +
  scale_y_log10() +
  labs(
    title = "Top 10 Most Abundant Metabolites by Treatment",
    x = "Metabolite",
    y = "Mean concentration across replicates (log10 scale)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    strip.text = element_text(face = "bold", size = 13),
    panel.spacing = unit(1.2, "lines")
  ) 
p_rank_top10_log

```

