---
title: "BIOL 3991_Omics Project"
author: "Thuy Le"
date: "2026-02-25"
output: html_document
---
# Import or install the packages to use later
```{r install package, message = FALSE}
library(readxl)
library(purrr)
library(tidyverse)
library(stringr)
library(viridis)
```

# Sync the excel file to OneDrive
If you haven't had OneDrive on your computer yet, download it and log in by mta email.

Go to your OneDrive on your computer, there is "BIOL 3991_Omics data" folder and our excel is in it. Log in to the OneDrive on your computer and make sure you download the folder in your OneDrive by press on the clound with down arrow icon next to the folder

This is the OneDrive folder link in case you missed it:
https://mountallison-my.sharepoint.com/:f:/r/personal/ttle_mta_ca/Documents/BIOL%203991_Omics%20data?csf=1&web=1&e=DoE83U

#Import datat
```{r import data, message=FALSE}
# Find your exact file path (erase the "#" at the start of the line to use it)
## Mac OS version
hits <- list.files(path.expand("~/Library/CloudStorage"),
                   pattern = "prodatawithCELLCOUNTS\\.xlsx$",
                   recursive = TRUE,
                   full.names = TRUE)
## Window version
#hits<-list.files(path.expand("~"),
#                 pattern="prodatawithCELLCOUNTS\\.xlsx$",
#                 recursive=TRUE,
#                 full. Names=TRUE)

hits # You will see your file path here 

# Paste your file path in ""
xlsx_path <- path.expand("/Users/kyoshiinari/Library/CloudStorage/OneDrive-MountAllisonUniversity/BIOL 3991_Omics data/prodatawithCELLCOUNTS.xlsx")

sheets <- excel_sheets(xlsx_path)
raw_data <- setNames(map(sheets, ~ read_excel(xlsx_path, sheet = .x)), sheets)
print(raw_data)
```

## Figure


```{r}


# ---- 1) Pick the sheet that contains the metabolite table ----
# (This auto-picks the first sheet that has a "mtabName" column)
sheet_name <- names(raw_data)[map_lgl(raw_data, ~ "mtabName" %in% names(.x))][1]
df <- raw_data[[sheet_name]]

# ---- 2) Identify sample columns (all CMP032* columns) ----
sample_cols <- grep("^CMP032", names(df), value = TRUE)

# ---- 3) Long format + log10 transform (exactly: log10(conc + 1e-12)) ----
# Also: turn 0 into NA so it plots as blank (white)
long <- df %>%
  select(mtabName, all_of(sample_cols)) %>%
  pivot_longer(-mtabName, names_to = "sample", values_to = "conc") %>%
  mutate(
    conc = as.numeric(conc),
    conc = na_if(conc, 0),  # make zeros blank on heatmap
    sample_disp = sample %>%
      str_replace_all("\\.+", "_") %>%
      str_replace_all("__+", "_") %>%
      str_replace_all("_$", ""),
    Condition = str_extract(sample_disp, "repl|lown|lowp"),
    Replicate = str_extract(sample_disp, "(?<=_)[A-Za-z]$"),
    Condition = factor(Condition, levels = c("repl", "lown", "lowp")),
    Replicate = factor(Replicate, levels = c("A", "B", "C")),
    log10_conc = log10(conc + 1e-12)  # <- this is where raw becomes log10
  )

# Keep sample order: repl A/B/C, lown A/B/C, lowp A/B/C
sample_levels <- long %>%
  distinct(sample_disp, Condition, Replicate) %>%
  arrange(Condition, Replicate) %>%
  pull(sample_disp)

# Keep metabolite order like your table, but flipped so first row ends up at bottom
met_levels <- rev(unique(df$mtabName))

long <- long %>%
  mutate(
    sample_disp = factor(sample_disp, levels = sample_levels),
    mtabName = factor(mtabName, levels = met_levels)
  )

# Optional: set legend range similar to your screenshot (so the scale looks comparable)
lims <- range(long$log10_conc, na.rm = TRUE)
lims <- c(min(lims[1], -10), max(lims[2], -2.5))

```

```{r heat map}
# Plot
p <- ggplot(long, aes(x = sample_disp, y = mtabName, fill = log10_conc)) +
  geom_tile() +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_viridis_c(option = "F", direction = -1, 
    na.value = "white",
    begin = 0.20,
    limits = lims,
    breaks = c(-2.5, -5, -7.5, -10)
) +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Metabolite Concentrations Heatmap",
    x = "Sample",
    y = "Metabolites",
    fill = "log10(Concentration + 1e-12)"
  ) +
#  theme_void() +
  theme(
#   panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 16),

    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
    axis.text.y = element_text(size = 11),

    legend.title = element_text(size = 13),
    legend.text  = element_text(size = 12),

    plot.margin = margin(10, 15, 10, 10)
  )

p

ggsave("metabolite_heatmap_log10.png",p, width = 12, height = 8, dpi = 300)
```

# Figure 2
```{r}
# Calculate log2 fold change relative to the repl mean for each metabolite
fc_long <- long %>%
  group_by(mtabName) %>%
  mutate(
    repl_mean = mean(conc[Condition == "repl"], na.rm = TRUE),
    log2_fc = log2(conc / repl_mean),
    log2_fc = ifelse(is.finite(log2_fc), log2_fc, NA)
  ) %>%
  ungroup()

# Use a symmetric legend around 0
fc_lim <- max(abs(fc_long$log2_fc), na.rm = TRUE)

```

```{r}
# Plot
p_fc <- ggplot(fc_long, aes(x = sample_disp, y = mtabName, fill = log2_fc)) +
  geom_tile(width = 1, height = 1) +
  geom_vline(xintercept = c(3.5, 6.5), linewidth = 1, color = "black") +
  scale_fill_gradient2(
    low = "#f4b6c2",     # light pink for negative
    mid = "grey95",      # near 0
    high = "red",        # positive
    midpoint = 0,
    na.value = "white",
    limits = c(-fc_lim, fc_lim),
    breaks = c(-8, -4, 0, 4, 8),
    name = "Log2FC"
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(
    title = "Differential Metabolic Profiles",
    subtitle = "Log2 Fold Change relative to Control (repl) mean",
    x = "Samples",
    y = "Metabolites"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    plot.subtitle = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12),
    axis.text.y = element_text(size = 11),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    plot.margin = margin(10, 15, 10, 10)
  )

p_fc

ggsave("metabolite_heatmap_log2fc.png", p_fc, width = 12, height = 8, dpi = 300)
```



